# Conditional Expectation and Projection {#sec-Conditional-Expectation-Projection}

In this chapter I replicate the examples from Hansen's Chapter 2: Conditional Expectation and Projection. 

The R packages used in the chapter are **HansenData**. Make sure these packages are installed an loaded prior to running the code in this chapter.

```{r}
#| include: false
options(digits=3)
```

## Load and Prepare Data

The examples in Hansen Chapter 2 are based on a sub-set of the March, 2009 CPS data available in the package **HansenData** as the tibble `cps09mar`:

```{r}
library(HansenData)
head(cps09mar)
```

In Section 2.2, Bruce considers the distribution of the random variable *wage*
$$
F(y) = \Pr[wage \le y]
$$
where *wage* is defines as weekly earnings. I create the wage and log wage data using:
```{r}
# cps09mar$experience = cps09mar$age - cps09mar$education - 6
cps09mar$wage = cps09mar$earnings/(cps09mar$hours*cps09mar$week)
cps09mar$lwage = log(cps09mar$wage)
attach(cps09mar)
```

Attaching a data frame (or tibble) using `attach()` allows the variables in the data frame to be referenced by name (e.g., use `wage` instead of by `cps09mar$wage`). This will be convenient in the following analysis. 

## Distribution of Hourly Wages

Summary statistics for `wage` and `lwage` can be computed using the generic function `summary()`:

```{r}
summary(round(cbind(wage, lwage),digits=2))
```
The mean hourly wage is about \$24/hr and the interquartile range is [\$12.8, \$28.9]. For `wage` the median wage, \$19.2, is smaller than the mean wage suggesting a positive skewness in the distribution of wages. For `lwage` the mean and median are very close suggesting that the distribution is symmetric. 

Kernel density estimates (smoothed histograms, see Hansen Chapter xx) of `wage` and `lwage` can be computed using the `density()` function:

```{r}
wage.density = density(wage, bw=5)
lwage.density = density(lwage, bw=0.1)
```

The argument `bw` specifies the bandwidth of the kernel density estimate and controls the smoothness of the resulting estimate. The density estimates are show in the figures below, replicating Hansen Figure 2.1:

```{r}
# kernel density estimate of pdf

par(mfrow=c(1,2))
plot(wage.density, main="", xlim=c(0, 70), xlab="(a) Wage Density",
       lwd=2, col="blue")
plot(lwage.density, main="", xlab="(b) Log Wage Density", 
     lwd=2, xlim = c(0, 6), col="blue")
```

```{r}
#| echo: false
par(mfrow=c(1,1))
```

## Mean and Median Wages

```{r}
mean(wage)
median(wage)
```

```{r}
plot(wage.density, main="pdf of wage", xlim=c(0, 100), xlab="Dollars per Hour",
       lwd=2, col="blue")
abline(v=mean(wage), col="blue", lwd=2)
abline(v=median(wage), col="red", lwd=2)
legend("topright", legend=c("Mean", "Median"), lty=1, lwd=2, col=c("blue", "red"))
```

## Mean and Median Log Wages

```{r}
mean(lwage)
median(lwage)
```

```{r}
plot(lwage.density, main="pdf of log wage", xlim=c(1, 6), xlab="Log Dollars per Hour",
       lwd=2, col="blue")
abline(v=mean(lwage), col="blue", lwd=2)
abline(v=median(lwage), col="red", lwd=2)
legend("topright", legend=c("Mean", "Median"), lty=1, lwd=2, col=c("blue", "red"))
```

## Conditional Distributions for Men and Women

```{r}
lwageWomen.density = density(lwage[female==1], bw=0.1)
lwageMen.density = density(lwage[female==0], bw=0.1)
plot(lwageWomen.density, main="pdf of log(wage) for Men and Women", 
     xlab="Log Dollars per Hour", lwd=2, xlim = c(1, 6))
points(lwageMen.density, type="l", col="blue", lwd=2)
legend("topright", legend=c("Women", "Men"), lty="solid", lwd=2, col=c("black", "blue"))
```

```{r}
# mean log wages for men and women
mean(lwage[female==0])
mean(lwage[female==1])
# difference
mean(lwage[female==0]) - mean(lwage[female==1])

```

## Conditional Distributions for Men and Women by race

```{r}
# race = 1 = white, race = 2 = black
lwageWhiteWomen.density = density(lwage[female==1 & race==1], bw=0.1)
lwageWhiteMen.density = density(lwage[female==0 & race==1], bw=0.1)
lwageBlackWomen.density = density(lwage[female==1 & race==2], bw=0.1)
lwageBlackMen.density = density(lwage[female==0 & race==2], bw=0.1)
plot(lwageWhiteWomen.density, main="pdf of log(wage) for Men and Women by race", 
     xlab="Log Dollars per Hour", lwd=2, xlim = c(1, 6), ylim=c(0, 0.8))
points(lwageWhiteMen.density, type="l", col="blue", lwd=2)
points(lwageBlackWomen.density, type="l", col="green", lwd=2)
points(lwageBlackMen.density, type="l", col="red", lwd=2)
legend("topright", legend=c("White Women", "White Men", "Black Women", "Black Men"), 
       lty="solid", lwd=2, col=c("black", "blue", "green", "red"))
```

## Conditional Means for Men and Women by race

```{r}
# White males
mean(lwage[female==0 & race==1])
# White females
mean(lwage[female==1 & race==1])
# Black males
mean(lwage[female==0 & race==2])
# Black females
mean(lwage[female==1 & race==2])
```

## Conditional Mean Function

Compute conditional means of log wages for men and women conditional on education level

```{r}
elevels = sort(unique(education))
cmean.mat = matrix(0, length(elevels), 2)
colnames(cmean.mat) = c("White.Male", "White.Female")
rownames(cmean.mat) = as.character(elevels)

for (i in 1:length(elevels)) {
  cmean.mat[i, 1] = mean(lwage[female==0 & race==1 & education==elevels[i]])
  cmean.mat[i, 2] = mean(lwage[female==1 & race==1 & education==elevels[i]])
}
```

## Plot Conditional Mean Function

```{r}
plot(elevels, cmean.mat[, 1], type="b", pch=16, lwd=2, 
     ylim=c(1.75,4), ylab="Log Dollars per Hour", xlab="Years of Education")
points(elevels, cmean.mat[, 2], type="b", pch=16, lwd=2, col="blue")
legend("topleft", legend=c("Men", "Women"), lty="solid", pch=16, lwd=2, col=c("black", "blue"))
```
