# Conditional Expectation and Projection {#sec-Conditional-Expectation-Projection}

In this chapter I replicate the examples from Hansen's Chapter 2: Conditional Expectation and Projection. 

The R packages used in the chapter are **HansenData**. Make sure these packages are installed an loaded prior to running the code in this chapter.

```{r}
#| include: false
options(digits=3)
```

## Load and Prepare Data

The examples in Hansen Chapter 2 are based on a sub-set of the March, 2009 CPS data available in the package **HansenData** as the tibble `cps09mar`:

```{r}
library(HansenData)
head(cps09mar)
```

In Section 2.2, Bruce considers the marginal distributions of the random variables *wage* and *log wage*:
$$
F(y) = \Pr[wage \le y], ~ F(\ell) = \Pr[log(wage) \le \ell]
$$
where *wage* is defines as weekly earnings. I create the wage and log wage data using:
```{r}
# cps09mar$experience = cps09mar$age - cps09mar$education - 6
cps09mar$wage = cps09mar$earnings/(cps09mar$hours*cps09mar$week)
cps09mar$lwage = log(cps09mar$wage)
attach(cps09mar)
```

Attaching a data frame (or tibble) using `attach()` allows the variables in the data frame to be referenced by name (e.g., use `wage` instead of by `cps09mar$wage`). This will be convenient in the following analysis. 

## Distribution of Hourly Wages

The distribution of hourly wages can be visualized using a histogram. In R, histograms can be created using the `hist()` function:

```{r}
#| label: fig-Histogram-Wages
#| echo: true
#| fig-cap: Histograms of hourly wages and log wages.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
par(mfrow=c(1,2))
hist(wage, main="", freq=FALSE, col="cornflowerblue")
hist(lwage, main="", freq=FALSE, col="cornflowerblue")
```

The histograms show that the distribution of hourly wages is heavily skewed to the right, while the distribution of log wages is more symmetric. 

Kernel density estimates (i.e., smoothed histograms, see Hansen Chapter 19) of `wage` and `lwage` can be computed using the `density()` function:

```{r}
wage.density = density(wage, bw=5)
lwage.density = density(lwage, bw=0.1)
```

The argument `bw` specifies the bandwidth of the kernel density estimate and controls the smoothness of the resulting estimate. The density estimates are show in the figures below, replicating Hansen Figure 2.1:

```{r}
#| label: fig-Smoothed-Histogram-Wages
#| echo: true
#| fig-cap: Density of Wages and Log Wages.
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
par(mfrow=c(1,2))
plot(wage.density, main="", xlim=c(0, 70), xlab="(a) Wage Density",
       lwd=2, col="blue")
plot(lwage.density, main="", xlab="(b) Log Wage Density", 
     lwd=2, xlim = c(0, 6), col="blue")
```


### Summary statistics wages

Summary statistics for `wage` and `lwage` can be computed using the generic function `summary()`:

```{r}
summary(round(cbind(wage, lwage),digits=2))
```
The mean hourly wage is about \$24/hr and the interquartile range is [\$12.8, \$28.9]. For `wage` the median wage, \$19.2, is smaller than the mean wage suggesting a positive skewness in the distribution of wages. For `lwage` the mean and median are very close suggesting that the distribution is symmetric. 

The R functions `mean()` and `median()` can be used to compute the mean and median of a vector:

```{r}
ans = c(mean(wage), median(wage))
names(ans) = c("mean", "median")
ans
```

A convenient function to compute a collection of the most common summary statistics is the `describe()` function from the **psych** package:

```{r}
library(psych)
describe(cbind(wage, lwage))
```

## Conditional Expectation

In Section 2.3, Hansen considers the conditional distribution of log wages for men and women. To compute the conditional means of log wages for men and women, I use the `mean()` function with the logical conditions `female==0` for men and `female==1` for women: 

```{r}
# mean log wages for men and women
mean_men = mean(lwage[female==0])
mean_women = mean(lwage[female==1])
mean_difference = mean(lwage[female==0]) - mean(lwage[female==1])
ans = c(mean_men, mean_women, mean_difference)
names(ans) = c("Men", "Women", "Diff")
ans
```
The distribution of log wages for men and women can be visualized using kernel density estimates:

```{r}
#| label: fig-Smoothed-Histogram-by-Gender
#| echo: true
#| fig-cap: Hansen Figure 2.2 panel (a)
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
lwageWomen.density = density(lwage[female==1], bw=0.1)
lwageMen.density = density(lwage[female==0], bw=0.1)
plot(lwageWomen.density, main="pdf of log(wage) for Men and Women", 
     xlab="Log Dollars per Hour", lwd=2, xlim = c(1, 6))
points(lwageMen.density, type="l", col="blue", lwd=2)
legend("topright", legend=c("Women", "Men"), lty="solid", lwd=2, col=c("black", "blue"))
```

Here, you see that the distribution of log wages for men is shifted to the right relative to the distribution for women.

### Conditional Distributions for Men and Women by race

The means of log wages for men and women by race can be computed using the `mean()` function with appropriate logical conditions (`race==1` for White and `race==2` for Black):

```{r}
mean_white_male = mean(lwage[female==0 & race==1])
mean_white_female = mean(lwage[female==1 & race==1])
mean_black_male = mean(lwage[female==0 & race==2])
mean_black_female = mean(lwage[female==1 & race==2])
ans = rbind(c(mean_white_male, mean_white_female), c(mean_black_male, mean_black_female))
dimnames(ans) = list(c("White", "Black"), c("Men", "Women"))
ans
```
To compute the conditional densities by gender and race use:

```{r}
#| label: fig-Smoothed-Histogram-by-Gender-Race
#| echo: true
#| fig-cap: Hansen Figure 2.2 panel (b)
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
lwageWhiteWomen.density = density(lwage[female==1 & race==1], bw=0.1)
lwageWhiteMen.density = density(lwage[female==0 & race==1], bw=0.1)
lwageBlackWomen.density = density(lwage[female==1 & race==2], bw=0.1)
lwageBlackMen.density = density(lwage[female==0 & race==2], bw=0.1)
plot(lwageWhiteWomen.density, main="pdf of log(wage) for Men and Women by race", 
     xlab="Log Dollars per Hour", lwd=2, xlim = c(1, 5), ylim=c(0, 0.8))
points(lwageWhiteMen.density, type="l", col="blue", lwd=2)
points(lwageBlackWomen.density, type="l", col="green", lwd=2)
points(lwageBlackMen.density, type="l", col="red", lwd=2)
legend("topright", legend=c("White Women", "White Men", "Black Women", "Black Men"), 
       lty="solid", lwd=2, col=c("black", "blue", "green", "red"))
```

## Logs and Percentages

TBD

```{r}

```


## Conditional Expectation Function

Consider computing the conditional means of log wages for men and women conditional on all observed education levels:

```{r}
elevels = sort(unique(education))
cmean.mat = matrix(0, length(elevels), 2)
colnames(cmean.mat) = c("White.Male", "White.Female")
rownames(cmean.mat) = as.character(elevels)

for (i in 1:length(elevels)) {
  cmean.mat[i, 1] = mean(lwage[female==0 & race==1 & education==elevels[i]])
  cmean.mat[i, 2] = mean(lwage[female==1 & race==1 & education==elevels[i]])
}
cmean.mat
```
These conditional means as a function of education level can be visualized using a line plot:

```{r}
#| label: fig-Conditional-Mean-by-Education
#| echo: true
#| fig-cap: Hansen Figure 2.3
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
plot(elevels, cmean.mat[, 1], type="b", pch=16, lwd=2,
     ylim=c(1.75,4), ylab="Log Dollars per Hour", xlab="Years of Education")
points(elevels, cmean.mat[, 2], type="b", pch=16, lwd=2, col="blue")
legend("topleft", legend=c("Men", "Women"), lty="solid", pch=16, lwd=2, col=c("black", "blue"))
```

## Continuous Variables

TBD

## CEF Error

TBD

## Illustrations of Best Linear Predictor

TBD