# Algebra of Least Squares {#sec-Algebra-Least-Squares}

In this chapter I replicate the examples from Hansen chapter 3, Algebra of Least Squares.

The R packages used in the chapter are **HansenData**. Make sure these packages are installed an loaded prior to running the code in this chapter.

```{r}
#| include: false
options(digits=3)
```

## Solving for Least Squares with One Regressor

In Section 3.5, Hansen considers the simple regression model

$$
Y_i = X_i \beta + e_i,  i=1, \ldots n
$$ 
where $X_i \sim U(0,1)$ is the regressor, $\beta = 3$ is the coefficient, and $e_i \sim N(0,1)$ is the error term. Hansen simulates data from this model, with $n=20$, using

```{r}
n = 20
set.seed(3)
X = runif(n,0,1)
e = rnorm(n)
Y = X*3 + e
```

The simulated data is shown in @fig-Simple-Regression below:

```{r}
#| label: fig-Simple-Regression
#| echo: true
#| fig-cap: Simulated data from the simple regression model
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
plot(X, Y, xlab="X", ylab="Y", pch=16, col="cornflowerblue", cex=1.5)
```

The idea behind least squares estimation is to find the best fitting line to the observed scatterplot of data. With the simple regression model above the linear relationship between $Y$ and $X$ has and intercept of zero and slope $\beta$.   

The least squares estimator $\hat{\beta}$ is the value of $\beta$ that minimizes the sum of squared errors (SSE):

The simple least squares estimator that minimizes $\mathrm{SSE}(\beta)=\sum_{i=1}^n (Y_i - X_i \beta)^2$ is 
$$
\hat{\beta} = \frac{\sum_{i=1}^n X_i Y_i}{\sum_{i=1}^n X_i^2}.
$$ 
This can be computed in R using^[The solution $\hat{\beta}=3.07$ given by Hansen on page 64 is the least square estimator computed with X and an intercept.]

```{r}
beta_hat = sum(X*Y)/sum(X^2)
beta_hat
```

The fitted regression line is shown in @fig-Simple-Regression-Fit below:

```{r}
#| label: fig-Simple-Regression-Fit
#| echo: false
#| fig-cap: Fitted regression line from the simple regression model
#| out-width: 80%
#| fig-asp: 0.75
#| fig-align: center
#| fig-legend: Fitted regression line
#| fig-legend-pos: bottom
plot(X, Y, xlab="X", ylab="Y", pch=16, cex=1.5, col="cornflowerblue")
abline(a=0, b=beta_hat, lwd=2)
```

The function $\mathrm{SSE}(\beta)$ 

## Illustration of Simple Regression Using Matrix Algebra

The example Hansen Chapter 2 Section 7, uses the same CPS data set as in @sec-Conditional-Expectation-Projection:

```{r}
library(HansenData)
cps09mar$experience = cps09mar$age - cps09mar$education - 6
cps09mar$wage = cps09mar$earnings/(cps09mar$hours*cps09mar$week)
cps09mar$lwage = log(cps09mar$wage)
```

* Do matrix algebra calculations first to illustrate matrix math in R
* Talk about why you should never calculate the projection matrix
* Then show how to use the R function `lm()` to estimate the simple regression model

First, I replicate the illustration of the regression of log wages on years of education and a constant (intercept) or married (spouse present) black female wage earners with 12 years of potential work experience:

$$
lnwage_{i} = \beta \times educ_{i} + \alpha + e_{i} = \mathrm{X}_i'\boldsymbol{\beta} + e_i, ~ i=1, \ldots n
$$
where $X_i = (educ_i, 1)'$ is the $(2 \times 1)$ vector of regressors, $\boldsymbol{\beta} = (\beta, \alpha)'$ is the vector of coefficients, and $e_i$ is the error term.

I construct the sample data using:
```{r}
exampleData = subset(cps09mar, subset = (race == 2 & female == 1 & marital == 1 & experience == 12))
head(exampleData)
```
There are `r nrow(exampleData)` observations in the sample.

In matrix notation, the regression model can be written as
$$
\underset{(n \times 1)}{\mathbf{Y}} = \underset{(n \times 2)}{\mathbf{X}}\underset{(2 \times 1)}{\boldsymbol{\beta}} + \underset{(n \times 1)}{\mathbf{e}}.
$$
The least squares estimator of $\boldsymbol{\beta}$ is
$$
\hat{\boldsymbol{\beta}} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}.
$$

Using the matrix algebra functions in R, the least squares estimator can be computed as follows:

```{r}
X = cbind(exampleData$education, 1)
Y = exampleData$lwage
beta_hat = solve(t(X) %*% X) %*% t(X) %*% Y
rownames(beta_hat) = c("education", "intercept")
colnames(beta_hat) = "estimate"
beta_hat
```
In R, `t()` transposes a matrix, `solve()` computes the inverse of a matrix, and `%*%` is the matrix multiplication operator.

For black females with 12 and 16 years of education, the predicted log wages are
```{r}
Y_hat_12 = beta_hat[1] * 12 + beta_hat[2]
Y_hat_16 = beta_hat[1] * 16 + beta_hat[2]
ans = c(Y_hat_12, Y_hat_16)
ans
```


### Fitted Values and Residuals

The $n \times 1$ vector of least squares fitted values are
$$
\hat{\mathbf{Y}} = \mathbf{X}\hat{\boldsymbol{\beta}} = \mathbf{X}(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}=\mathbf{P}\mathbf{Y},
$$
where $\mathbf{P} = \mathbf{X}(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'$ is the $n \times n$ projection matrix. Recall, $\mathbf{P}$ is a symmetric and idempotent matrix satisfying $\mathbf{P}^2 = \mathbf{P}\mathbf{P} = \mathbf{P}$ and $\mathbf{P}' = \mathbf{P}$.

The $n \times 1$ vector of least squares residuals are
$$
\hat{\mathbf{e}} = \mathbf{Y} - \hat{\mathbf{Y}} = \mathbf{Y} - \mathbf{P}\mathbf{Y} = (\mathbf{I} - \mathbf{P})\mathbf{Y} = \mathbf{M}\mathbf{Y},
$$

where $\mathbf{M} = \mathbf{I} - \mathbf{P}$ is the idempotent annihilator matrix orthogonal to $\mathbf{P}$ satisfying $\mathbf{M}^2 = \mathbf{M}\mathbf{M} = \mathbf{M}$, $\mathbf{M}' = \mathbf{M}$ and $\mathbf{M}\mathbf{P} = \mathbf{0}$.

The fitted values and residuals can be computed in R as follows:

```{r}
Y_hat = X %*% beta_hat
e_hat = Y - Y_hat
ans = cbind(Y_hat, e_hat)
colnames(ans) = c("fitted values", "residuals")
head(ans)
```
Equivalently, the fitted values and residuals can be computed using the projection and annihilator matrices:

```{r}
P = X %*% solve(t(X) %*% X) %*% t(X)
M = diag(nrow(X)) - P
Y_hat = P %*% Y
e_hat = M %*% Y
ans = cbind(Y_hat, e_hat)
colnames(ans) = c("fitted values", "residuals")
head(ans)
```
You can verify numerically that $\mathbf{P}^2=\mathbf{P}$, $\mathbf{P}' = \mathbf{P}$, and $\mathbf{P}\mathbf{M} = \mathbf{0}$. 

In R, it is easy to verify the six conditions of Theorem 3.3. For example, to verify $tr(\mathbf{P})=k$ use
```{r}
sum(diag(P))
```
To verify $\mathbf{P}$ has k eigenvalues equal to 1 and $n-k$ eigenvalues equal to 0, use
```{r}
round(eigen(P)$values)
```

For computing the fitted values and residuals it is not recommended to use the projection and annihilator matrices because they are $n \times n$ matrices and can be very large if $n$ is large. Instead, use the formulas $\hat{\mathbf{Y}} = \mathbf{X}\hat{\boldsymbol{\beta}}$ and $\hat{\mathbf{e}} = \mathbf{Y} - \hat{\mathbf{Y}}$.

### Properties of Fitted Values and Residuals

Residuals have mean zero:

```{r}
round(mean(e_hat))
```
Residuals are uncorrelated with regressors (education and constant) and fitted values:

```{r}
round(as.numeric(cor(e_hat, exampleData$education)))
round(as.numeric(cor(e_hat, Y_hat)))
```
### Estimation of Error Variance

The error variance, $\sigma^2 = \mathrm{Var}(e_i)$, can be estimated using the residuals in the following two ways:
$$
\tilde{\sigma}^2 = \frac{1}{n - k} \sum_{i=1}^n \hat{e}_i^2 \quad \text{and} \quad \hat{\sigma}^2 = \frac{1}{n - k} \sum_{i=1}^n \hat{e}_i^2
$$
The estimate $\tilde{\sigma}^2$ is the maximum likelihood estimator of $\sigma^2$ and $\hat{\sigma}^2$ is the unbiased (degree-of-freedom adjusted) estimator of $\sigma^2$. In R, these estimates can be computed as follows:

```{r}
n.obs = nrow(exampleData)
sigma2_tilde = sum(e_hat^2)/(n.obs - 2)
sigma2_hat = sum(e_hat^2)/n.obs
c(sigma2_tilde, sigma2_hat)

```
### Analysis of Variance and Regression $R^2$

Using the trivial identity
$$
\mathbf{Y} = \mathbf{\hat{Y}} + \mathbf{Y} - \mathbf{\hat{Y}} = \mathbf{\hat{Y}} + \mathbf{\hat{e}}
$$
and the fact that $\mathbf{\hat{Y}}$ and $\mathbf{\hat{e}}$ are orthogonal gives the sum of squares decomposition
$$
\mathbf{Y}'\mathbf{Y} = \mathbf{\hat{Y}}'\mathbf{\hat{Y}} + \mathbf{\hat{e}}'\mathbf{\hat{e}}.
$$

If a constant is in the regression, then the sum of squares decomposition can be expressed as
$$
(\mathrm{\mathbf{Y}} - \mathrm{1}\bar{Y})'(\mathrm{\mathbf{Y}} - \mathrm{1}\bar{Y}) = (\mathbf{\hat{Y}} - \mathrm{1}\bar{Y})'(\mathbf{\hat{Y}} - \mathrm{1}\bar{Y}) + \mathbf{\hat{e}}'\mathbf{\hat{e}},
$$
or
$$
\sum_{i=1}^n (Y_i - \bar{Y})^2 = \sum_{i=1}^n (\hat{Y}_i - \bar{Y})^2 + \sum_{i=1}^n \hat{e}_i^2
$$
Dividing both sides by $n$ gives rise to the variance decomposition:
$$
\widehat{\mathrm{var}}(\mathbf{Y}) = \widehat{\mathrm{var}}(\mathbf{\hat{Y}}) + \widehat{\mathrm{var}}(\mathbf{\hat{e}}).
$$

Further dividing both sides by $\widehat{\mathrm{var}}(\mathbf{Y})$ gives the $R^2$ decomposition:
$$
1 = \frac{\widehat{\mathrm{var}}(\mathbf{\hat{Y}})}{\widehat{\mathrm{var}}(\mathbf{Y})} + \frac{\widehat{\mathrm{var}}(\mathbf{\hat{e}})}{\widehat{\mathrm{var}}(\mathbf{Y})} = R^2 + (1 - R^2).
$$
Here $R^2 = \widehat{\mathrm{var}}(\mathbf{\hat{Y}})/\widehat{\mathrm{var}}(\mathbf{Y})$ is the proportion of the variance in $\mathbf{Y}$ explained by the regression, and $1 - R^2 = \widehat{\mathrm{var}}(\mathbf{\hat{e}})/\widehat{\mathrm{var}}(\mathbf{Y})$ is the proportion of the variance in $\mathbf{Y}$ not explained by the regression.

## Illustration of Simple Regression Using `lm()`

Use the R function `lm()` for least squares estimation

```{r}
lwage.fit = lm(lwage ~ education, data=exampleData)
class(lwage.fit)
names(lwage.fit)
```

The component `coefficients` contains the estimated coefficients $\hat{\beta}$, `fitted.values` contains the fitted values $\hat{\mathbf{Y}}$, and `residuals` contains the residuals $\hat{\mathbf{e}}$.

* Disscuss the other components
* Discuss the qr decomposition used in the `lm()` function to compute the least squares estimates. The `qr` component contains the QR decomposition of the regressor matrix $\mathbf{X}$, which is used to compute the least squares estimates efficiently and numerically stable.



Individual components can be extracted using `$`

```{r}
lwage.fit$coefficients
head(lwage.fit$fitted.values)
head(lwage.fit$residuals)
```

Coefficients, fitted values, and residuals can also be extracted using the extractor functions `coef()`, `fitted()`, and `residuals()`:

```{r}
coef(lwage.fit)
head(fitted(lwage.fit))
head(residuals(lwage.fit))
```

R is object oriented and there are generic method functions `print()`, `summary()`, `plot()`, and `predict()`:

The `print` method shows the basic output from the estimation:

```{r}
lwage.fit
```

The `summary` method shows standard errors, t-statistics and other aspects of the fit

```{r}
summary(lwage.fit)
```

I will discuss the `summary`, `plot` and `predict` methods later on after we cover the material in chapter 4.

## Visualization of the Least Squares Fit

The least squares fit from a simple regression can be visualized in a scatterplot:

```{r}
plot(exampleData$education, exampleData$lwage, xlab="Years of education", 
     ylab="Log hourly wage", pch=16, col="cornflowerblue", cex=2)
abline(lwage.fit, lwd=2)
```


## Estimation of error variance

```{r}
n.obs = nrow(exampleData)
sig2hat = sum(residuals(lwage.fit)^2)/(n.obs - 2)
sighat = sqrt(sig2hat)
sig2hat
sighat
```

## Influence Statistics

The diagonal elements of the "hat" or projection matrix (the leverage points), can be computed using the R function `hatvalues()`:

```{r}
h = hatvalues(lwage.fit)
h
```

Here there are two regressors so the sum of the leverage points is equal to 2

```{r}
sum(h)
```

The R function `influence()` computes the leverage points as well as the change in the estimated coefficients which results when the i-th case is dropped from the regression: $\hat{\beta} - \hat{\beta}_{(-i)}$

```{r}
leverage.vals = influence(lwage.fit)
names(leverage.vals)
# change in the estimated coefficients which results when the i-th case is dropped from the regression
leverage.vals$coefficients
```

## Prediction errors

The vector of prediction errors is easy to compute:

```{r}
n.obs = nrow(exampleData)
Mstar = solve(diag(n.obs) - diag(h))
predError = as.numeric(Mstar%*%residuals(lwage.fit))
```

The sample mean squared prediction error (aka *cross validation statistic*) is

```{r}
cvStat = as.numeric(crossprod(predError)/n.obs)
cvStat
```

## Influence statistic

The influence values (diference between OLS fitted values and leave-one-out predicted values) can be computed using

```{r}
influence.vals = h*predError
influence.vals
influence.stat = max(influence.vals)
influence.stat
```

## Multiple Regression

Here I replicate the illustration of multiple regression from Hansen section 3.7. The multiple regression is

$$lnwage_{i} = \beta_{1} \times educ_{i} + \beta_{2} \times exper_{i} + \beta_{3} \times exper_{i}^{2}/100 +  \alpha + e_{i}$$

for single (never married) asian male wage earners

```{r}
exampleData2 = subset(cps09mar, subset = (race == 4 & female == 0 & marital == 7))
nrow(exampleData2)
```

Use the R function `lm()` for least squares estimation

```{r}
lwage.fit2 = lm(lwage ~ education + experience + I(experience^2 / 100), data=exampleData2)
class(lwage.fit2)
names(lwage.fit2)
```

```{r}
lwage.fit2 
```
